---
title: "Exercises"
output: 
  html_document: 
    toc: yes
---

## Chapter 2 - Getting started with ggplot2

```{r}
require(ggplot2, quietly = TRUE)
require(dplyr, quietly = TRUE, warn.conflicts = FALSE)
options(digits = 4, dplyr.print_min = 6, dplyr.print_max = 6)
mpg
```

### Exercises 2.2.1 (Page 14 f.)

__Question 1__: List five functions that you could use to get more information
    about the `mpg` dataset.  
__Answer__: Some useful functions are

 + `summary`: print some summary statistics for each variable  
 + `View`: to see the whole dataset in a nice spread-sheet like fashion
 + `str` : get info on the structure of the `mpg` object
 + `dplyr::glimps`: similar to `str` but much tidyer
 + `class`: to get its class
 
```{r}
glimpse(mpg)
```

__Question 2__: How can you find out what other datasets are included with ggplot2?  
__Answer__: You can find a list of all data set included in ggplot2 at <http://docs.ggplot2.org/current/index.html>.
    
__Question 3__: Apart from the US, most countries use fuel consumption (fuel consumed
    over fixed distance) rather than fuel economy (distance travelled with
    fixed amount of fuel). How could you convert `cty` and `hwy` into the
    European standard of l/100km?  
__Answer__:  

```{r}
mpgTol100km <- function(milespergallon){
  GalloLiter <- 3.785411784
  MileKilometer <- 1.609344 
  
  l100km <- (100*GalloLiter)/(milespergallon*MileKilometer)
  l100km
}
```
We could use `apply` to convert the columns
```{r}
apply(mpg[, c("cty", "hwy")], 2, mpgTol100km) %>% 
  head()
```

Another possibility is to use the `dplyr` functions `transmute` or `mutate` 
```{r}
mpg %>% 
  transmute("cty_l100km" = mpgTol100km(cty),
            "hwy_l100km" = mpgTol100km(hwy))
```

__Question 4__: Which manufacturer has the most models in this dataset? Which model has 
    the most variations? Does your answer change if you remove the redundant
    specification of drive train (e.g. "pathfinder 4wd", "a4 quattro") from the 
    model name?  
__Answer to the first part__: 
1.  Most models in total by manufacturer:
```{r}
q4_1 <- mpg %>% 
  group_by(manufacturer) %>% 
  tally(sort = TRUE)
```
1.1 If we are looking for most *unique* models by manufacturer 
```{r}
q4_2 <- mpg %>% 
  group_by(manufacturer) %>% 
  transmute("n" = length(unique(model))) %>% 
  unique() %>%
  ungroup() %>% 
  arrange(desc(n))
```

All combined in a plot

```{r}

df <- data.frame(rbind(q4_1, q4_2), "Type" = as.factor(rep(c("overall", "unique"), each = 15)))

ggplot(df, aes(x = reorder(as.factor(manufacturer), n), y = n, fill = Type)) + 
  geom_bar(width = 0.2, stat = "identity", position = "dodge") + 
  coord_flip() + 
  ggtitle("Number of models per manufacturer") +
  ylab("Count") + xlab("Model")
```

__Answer to the second part__

table(mpg$model)

#### Answers